---
title: "Marketing Spend and Customer Purchase Amount"
author: "Rikhitha Varma"
format:
  html:
    theme: cosmo
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
    documentclass: article
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(lubridate)
library(knitr)
library(ggplot2)

```


```{r}
campaign_raw <- read_csv("data/Marketing_campaign.csv")
```


```{r}
campaign_cleanf <- campaign_raw |>
  mutate(
    signup_date = parse_date_time(
      signup_date,
      orders = c("ymd", "dmy", "mdy", "b d, Y", "B d, Y"),
      quiet = TRUE
    )
  ) |>
  filter(!is.na(signup_date), !is.na(purchase_amount))
```

# Introduction

Marketing campaigns often vary the level of investment allocated to individual customers in order to improve sales outcomes. Understanding whether higher marketing spend is associated with greater customer purchases is important for improving the effectiveness of campaign strategies. This report analyses customer-level data from a personalised marketing campaign conducted by an online retailer. The aim is to evaluate whether higher marketing spend per customer is associated with higher purchase amounts.

# Research Question

This report investigates whether higher marketing spend per customer leads to higher purchase amounts. The analysis examines the relationship between marketing spend and purchase amount using customer-level campaign data. It also considers whether this relationship differs across customer segments.

# Dataset Introduction

The dataset contains customer-level information from a personalised marketing campaign conducted over one quarter. Each observation represents a unique customer and includes variables such as marketing spend, number of visits, purchase amount, customer segment, and signup date. To prepare the data for analysis, the signup date variable was converted into a standard date format using multiple parsing methods. Observations with missing values in signup date and purchase amount were removed to improve data quality. The cleaned dataset was then used for descriptive analysis and visualisation.

## Dataset Description

The cleaned dataset contains `r nrow(campaign_cleanf)` observations and `r ncol(campaign_cleanf)` variables. The dataset includes numeric variables such as marketing spend, visits, and purchase amount, categorical variables such as customer segment and notes, and a date variable representing customer signup date.


```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for the cleaned campaign dataset."

summary_table <- campaign_cleanf |>
  summarise(
    observations = n(),
    mean_marketing_spend = mean(marketing_spend, na.rm = TRUE),
    median_marketing_spend = median(marketing_spend, na.rm = TRUE),
    mean_visits = mean(visits, na.rm = TRUE),
    mean_purchase_amount = mean(purchase_amount, na.rm = TRUE),
    median_purchase_amount = median(purchase_amount, na.rm = TRUE)
  )

kable(
  summary_table,
  digits = 2
)
```

Table @tbl-summary presents summary statistics for the key variables in the dataset. The results show that both marketing spend and purchase amount vary across customers, indicating diversity in customer behaviour and spending patterns. The average values provide an overall view of campaign performance, while the median values help account for possible skewness in the data. Overall, these statistics provide a baseline understanding of customer activity and purchasing behaviour.


```{r}
#| label: tbl-segment
#| tbl-cap: "Average marketing spend and purchase amount by customer segment."

campaign_cleanf |>
  group_by(segment) |>
  summarise(
    avg_marketing_spend = mean(marketing_spend, na.rm = TRUE),
    avg_purchase_amount = mean(purchase_amount, na.rm = TRUE)
  ) |>
  kable(digits = 2)
```

Table @tbl-segment shows the average marketing spend and purchase amount across customer segments. The results indicate that customer behaviour differs between segments, with some groups receiving higher marketing investment and generating higher purchase amounts than others.


```{r}
#| label: fig-segment-summary
#| fig-cap: "Average purchase amount by customer segment."

campaign_cleanf |>
  group_by(segment) |>
  summarise(avg_purchase = mean(purchase_amount, na.rm = TRUE)) |>
  ggplot(aes(x = segment, y = avg_purchase)) +
  geom_col() +
  labs(
    x = "Customer segment",
    y = "Average purchase amount"
  )
```

Figure @fig-segment-summary shows the average purchase amount by customer segment. The figure indicates that customer purchasing behaviour differs across segments, with some segments generating higher average purchase amounts than others.

# Results


```{r}
#| label: fig-spend-purchase
#| fig-cap: "Relationship between marketing spend and purchase amount."

ggplot(campaign_cleanf, aes(x = marketing_spend, y = purchase_amount)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Marketing spend per customer",
    y = "Purchase amount"
  )
```


```{r}
#| label: fig-spend-segment
#| fig-cap: "Relationship between marketing spend and purchase amount across customer segments."

ggplot(campaign_cleanf, aes(x = marketing_spend, y = purchase_amount, colour = segment)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Marketing spend per customer",
    y = "Purchase amount",
    colour = "Segment"
  )
```

Figure @fig-spend-purchase shows the relationship between marketing spend and purchase amount. The scatter plot and fitted regression line indicate a clear positive association between the two variables. As marketing spend per customer increases, purchase amount also tends to increase. This suggests that customers who received higher levels of marketing investment generally made larger purchases.

Figure @fig-spend-segment further explores this relationship across customer segments. A positive trend is visible within all three segments, although the strength of the relationship varies. The Enterprise segment appears to show a steeper increase, while the Consumer and SMB segments display a more moderate upward trend. Overall, the results suggest that higher marketing spend is associated with higher purchase amounts, although the effectiveness of marketing investment differs across customer groups.

# Conclusion

This report analysed the relationship between marketing spend and customer purchase amount using campaign data. The results show a clear positive association, indicating that higher marketing spend is generally linked to higher purchase amounts. While this relationship is visible across customer segments, the strength of the pattern varies between groups. These findings suggest that increasing marketing investment can improve sales outcomes, although segment-specific strategies may further enhance campaign effectiveness.